---
title: "COMPSCIX 415.2 Homework 1"
output: html_document
author: Sanatan Das
date : 02/05/2018
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
My Github repository for my assignments can be found at this URL:https://github.com/sanatanonline/compscix-415-2-assignments

```{r load_packages, warning=FALSE, message=FALSE}
library(mdsr)
library(tidyverse)
data("WorldCities")
```

QUESTION 1: How many observations and variables are there in this data set? What are
some of the variable names? Type up your answer in your document using complete sentences.

ANSWER 1: There are 23018 observations and 10 variables in the data set. The variables are code, name etc. Please see below.

```{r glimpse WorldCities}
glimpse(WorldCities)
```

```{r extract part of the WorldCities data}
WorldCities <- head(WorldCities, 200) # 200 rows
country_col <- WorldCities$country
unique(country_col)
```

QUESTION 2: There is a column called region in the data set. Can you extract this and show
only the unique values?

AnSWER 2: 

```{r extract unique regions}
region = WorldCities$region
unique(region)
```


QUESTION 3: Can you extract and show the unique entries from the country column in
WorldCities using one line of code and two %>% operators?

ANSWER 3:

```{r extract unique countries}
WorldCities %>% select(country) %>% unique()
```

QUESTION 4: Make one more histogram of the population and add a new x-label, a new ylabel,
a new title (use ggtitle('my title')), and change the theme of the plot using theme_bw().

ANSWER 4:

```{r city population histogram}
WorldCities %>% ggplot(aes(x = population)) +
geom_histogram() +
xlab('City Population') + 
ylab('Population Count') + 
ggtitle('City Population Histogram') + 
theme_bw()
```


