---
title: "COMPSCIX 415.2 Homework 9/Final"
author: "Sanatan Das"
date: "March 29, 2018"
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{center}\LARGE\includegraphics[width=5cm]{logo.png}\\[\bigskipamount]}
  - \posttitle{\end{center}}
output:
  pdf_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage


# Code and Documents Git Repository

All the work can be found in the below Git repository location:

https://github.com/sanatanonline/compscix-415-2-assignments


# Load packages (prerequisites to run the code in this document)
```{r load_packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(rpart)
library(partykit)
library(modelr)
library(randomForest)
```

# Bootstrapping (10 points)

## Load the Titanic dataset

**1. Follow these steps:**

* **Load the train.csv dataset into R.**

* **Convert all character columns into unordered factors.**

* **Convert the Survived column into an unordered factor because it is loaded as an integer by default.**

* **Take a glimpse of your data to confirm that all of the columns were converted correctly.**

**We will use this same dataset for this entire assignment.**


**Answer:**

```{r var1, eval=TRUE, echo=TRUE}

# load train.csv
train = read_csv("C:/view/opt/apps/git/compscix-415-2-assignments/titanic/train.csv")
# glimpse train
glimpse(train) 
# Add the factors
train$Survived <- as.factor(train$Survived)
train$Name <- as.factor(train$Name)
train$Sex <- as.factor(train$Sex)
train$Ticket <- as.factor(train$Ticket)
train$Cabin <- as.factor(train$Cabin)
train$Embarked <- as.factor(train$Embarked)
# glimpse train again
glimpse(train)

```

Now all the character columns are converted to unordered factors.

## Bootstrap samples (100) of titanic data

**2. Use the code below to take 100 bootstrap samples of your data. Confirm that the result is a tibble with a list column of resample objects - each resample object is a bootstrap sample of the titanic dataset.**

```{r var2, eval=FALSE, echo=TRUE}

library(tidyverse)
library(modelr)
titanic_boot <- bootstrap(data = ____, n = ___)

```

**Answer:**

Let's fill in the code.

```{r var3, eval=TRUE, echo=TRUE}

titanic_boot <- bootstrap(data = train, n = 100)
#see the type of titanic_boot
class(titanic_boot)
# see titanic_boot
titanic_boot

```


